---
title: "Rmarkdown Final Project"
author: "Jordyn Blauer"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ABSTRACT


# BACKGROUND


# STUDY QUESTION and HYPOTHESIS

## Questions

Our study question is: "Does air quality have an effect on the amount of deaths caused by asthma?"

## Hypothesis

Our hypothesis is: "California will be the worst for those with asthma because it has the highest percentage of bad air quality."

A possible visualization we could use for our data is a barplot.

## Prediction



# METHODS


## Fill in 1st analysis 
```{r}
Group=read.csv("Group Project CompBio - Sheet1 (1).csv")
```
##gallery code
```{r, eval= FALSE}
# Load ggplot2
library(ggplot2)

# Create data
data <- data.frame(
  name=c("A","B","C","D","E") ,  
  value=c(3,12,5,18,45)
  )

# Barplot
ggplot(data, aes(x=name, y=value)) + 
  geom_bar(stat = "identity")
```

##gallery code with our data
```{r eval=FALSE}
# library
library(ggplot2)
 
# Load the data
air_data <- data.frame(State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"))
```

```{r}
# Load ggplot2
library(ggplot2)

# Create data
data <- data.frame(
  Asthma = c(9.6, 10.7, 9.7, 10.5, 8.7, 10.8, 12.4, 9.9, 9.3, 9.6, 9.1, 11.1, 8.7, 11, 9.7, 10.7, 10.8, 10, 13.1, 10.4, 11.3, 11.9, 9.9, 9.4, 10.4, 11.7, 8.1, 10.1, 13.1, 8.9, 10.4, 10.3, 9.2, 10.4, 11.4, 12.3, 11.5, 10.1, 13.3, 9, 8.3, 11.7, 7.9, 11, 12.9, 9.9, 10.9, 12.9, 10.9, 10.5), 
  Poor_Air = c(0.27, 0.12, 0.33, 0.3, 0.46, 0.25, 0.22, 0.31, 0.24, 0.3, 0.03, 0.28, 0.34, 0.25, 0.27, 0.29, 0.23, 0.29, 0.08, 0.19, 0.23, 0.24, 0.14, 0.31, 0.22, 0.19, 0.1, 0.25, 0.1, 0.27, 0.22, 0.16, 0.2, 0.17, 0.25, 0.34, 0.29, 0.31, 0.16, 0.25, 0.15, 0.23, 0.33, 0.29, 0.13, 0.12, 0.24, 0.16, 0.2, 0.15))

State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")

# Barplot
ggplot(data, aes(x=Poor_Air, y=State)) + 
  geom_bar(stat = "identity")
```
```{r asthma-ranked-bars, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=12, fig.cap="Adults with Current Asthma vs. Poor-Air Days by State (ranked by Asthma %)"}
# Use the existing air_data from our environment
# Convert to appropriate format and calculate percentages
df <- data.frame(
  State = air_data$State,
  asthma_pct = air_data$Asthma,
  poor_air_pct = air_data$Poor_Air * 100  # Convert proportion to percentage
)

# Sort by asthma percentage (descending)
df <- df[order(-df$asthma_pct), ]

# Set up plotting parameters
par(mar = c(5, 10, 4, 6))  # Adjust margins for state names
y_pos <- 1:nrow(df)  # Y positions for bars

# Create empty plot
plot(0, 0, type = "n", 
     xlim = c(0, 50), 
     ylim = c(0.5, nrow(df) + 0.5),
     xlab = "Percent", 
     ylab = "",
     main = "Adults with Current Asthma vs. Poor-Air Days by State\n(Ranked by Asthma %)",
     yaxt = "n")  # No default y-axis

# Add asthma bars (blue)
barplot_height <- df$asthma_pct
rect(xleft = 0, ybottom = y_pos - 0.3, 
     xright = barplot_height, ytop = y_pos + 0.3,
     col = "lightblue", border = NA)

# Add air quality bars (red) - offset slightly
barplot_height_air <- df$poor_air_pct
rect(xleft = 0, ybottom = y_pos - 0.3 + 0.6, 
     xright = barplot_height_air, ytop = y_pos + 0.3 + 0.6,
     col = "lightcoral", border = NA)

# Add state names
axis(2, at = y_pos, labels = df$State, las = 1, cex.axis = 0.6)

# Add legend
legend("bottomright", 
       legend = c("Adults with Current Asthma (%)", "Poor Air Quality Days (%)"),
       fill = c("lightblue", "lightcoral"),
       cex = 0.8,
       bty = "n")
```



```{r scatter plot}
# Scatter plot with regression line and statistics
plot(air_data$Poor_Air, air_data$Asthma,
     xlab = "Proportion of Poor Air Quality Days",
     ylab = "Asthma Rate (%)",
     main = "Air Quality vs Asthma Rates by State",
     pch = 16, col = "blue", cex = 1.2)

# Add regression line
model <- lm(Asthma ~ Poor_Air, data = air_data)
abline(model, col = "red", lwd = 2)

# Calculate statistics
correlation <- cor(air_data$Poor_Air, air_data$Asthma)
r_squared <- summary(model)$r.squared
p_value <- summary(model)$coefficients[2,4]

# Add statistics to plot
legend("topright", 
       legend = c(paste("Correlation:", round(correlation, 3)),
                  paste("R-squared:", round(r_squared, 3)),
                  paste("p-value:", round(p_value, 4))),
       bty = "n")

# Print statistics to console
cat("=== REGRESSION STATISTICS ===\n")
cat("Correlation:", round(correlation, 3), "\n")
cat("R-squared:", round(r_squared, 3), "\n")
cat("p-value:", round(p_value, 4), "\n")
cat("Regression equation: Asthma =", 
    round(coef(model)[1], 2), "+", 
    round(coef(model)[2], 2), "* Poor_Air\n")
```



# Second Analysis

## Fill in 2nd analysis/plot



## Statisical Test Code

\`\`\`


# DISCUSSION

## Interpretation of 1st analysis


## Interpretation of 2nd analysis 



# REFERENCES

