---
title: "Rmarkdown Final Project"
author: "Jordyn Blauer"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ABSTRACT


# BACKGROUND


# STUDY QUESTION and HYPOTHESIS

## Questions

Our study question is: "Does air quality have an effect on the amount of deaths caused by asthma?"

## Hypothesis

Our hypothesis is: "California will be the worst for those with asthma because it has the highest percentage of bad air quality."

A possible visualization we could use for our data is a scatterplot.

## Prediction



# METHODS


## Fill in 1st analysis 
```{r}
Group=read.csv("Group Project CompBio - Sheet1.csv")
```
##gallery code
```{r}
# library
library(ggplot2)
 
# The iris dataset is provided natively by R
#head(iris)
 
# basic scatterplot
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width)) + 
    geom_point()
```

##gallery code with our data
```{r eval=FALSE}
# library
library(ggplot2)
 
# basic scatterplot
ggplot(Group, aes(x=Average.proportion.per.state, y=State)) + 
    geom_point()

```


  
## Barplot Code




# Second Analysis

## Fill in 2nd analysis/plot



## Statisical Test Code

\`\`\`


# DISCUSSION

## Interpretation of 1st analysis


## Interpretation of 2nd analysis 



# REFERENCES

